---
title: "Assignment 2- Spatial Analysis"
author: "Sydney Rilum"
date: "10/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(stars)
library(stringr)
library(rgdal)
```

## Summary

“In February 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10–11, 13–17, and 15–20.”

In the project, we aim to:

- Estimate how many homes in Houston went without power as a result of the first two storms.
- Investigate the differences in recovery of different suburbs in Houston. Specifically, investigate if any socioeconomic factors are predictors of a community’s recovery from a power outage.

## Load the night lights image data

Night lights image data for clear days before and after the storm (2021-02-07, 2021-02-16) were acquired from NASA’s Level-1 and Atmosphere Archive & Distribution System Distributed Active Archive Center (LAADS DAAC). 

```{r}
# Create a function to load the DNB dataset from VNP46A1 granules

read_dnb <- function(file_name) {

  # Reads the "DNB_At_Sensor_Radiance_500m" dataset from a VNP46A1 granule into a STARS object.
  # Then read the sinusoidal tile x/y positions and adjust the STARS dimensions (extent+delta)

  # the HDF dataset (within the file) that contains the night lights band
  dataset_name <- "//HDFEOS/GRIDS/VNP_Grid_DNB/Data_Fields/DNB_At_Sensor_Radiance_500m"

  # Extract strings from file metadata containing the horizontal and vertical tile coords
  h_string <- gdal_metadata(file_name)[199]
  v_string <- gdal_metadata(file_name)[219]
  
  # Parse the h/v string to extract the integer number of h and v coords
  tile_h <- as.integer(str_split(h_string, "=", simplify = TRUE)[[2]])
  tile_v <- as.integer(str_split(v_string, "=", simplify = TRUE)[[2]])

  # Calculate geographic bounding box from tile coords (tile is 10 x 10 degrees)
  west <- (10 * tile_h) - 180
  north <- 90 - (10 * tile_v)
  east <- west + 10
  south <- north - 10

  # Calculate cell size in geographic coords (a tile is 10 degrees and has 2400 x 2400 grid cells/pixels)
  # NOTE: this is just to determine the coords of each pixel; we're NOT going to measure distances or areas in degrees! 
  delta <- 10 / 2400

  # Read the dataset
  dnb <- read_stars(file_name, sub = dataset_name)

  # Set CRS, apply origin (aka offset) and cell size (aka delta)
  st_crs(dnb) <- st_crs(4326)
  st_dimensions(dnb)$x$delta <- delta
  st_dimensions(dnb)$x$offset <- west
  st_dimensions(dnb)$y$delta <- -delta
  st_dimensions(dnb)$y$offset <- north
  
  return(dnb)
}
```

```{r}
# Call the read_dnb() four times to load the four datasets into four stars objects:

day38_tile85_2021039064328 <- "data/VNP46A1.A2021038.h08v05.001.2021039064328.h5"
dnb_day38_tile85_2021039064328 <- read_dnb(file_name = day38_tile85_2021039064328)

day38_tile86_2021039064329 <- "data/VNP46A1.A2021038.h08v06.001.2021039064329.h5"
dnb_day38_tile86_2021039064329 <- read_dnb(file_name = day38_tile86_2021039064329)

day47_tile85_2021048091106 <- "data/VNP46A1.A2021047.h08v05.001.2021048091106.h5"
dnb_day47_tile85_2021048091106 <- read_dnb(file_name = day47_tile85_2021048091106)

day47_tile86_2021048091105 <- "data/VNP46A1.A2021047.h08v06.001.2021048091105.h5"
dnb_day47_tile86_2021048091105 <- read_dnb(file_name = day47_tile86_2021048091105)
```


## Combine the data

```{r}
# Combine each pair of tiles into a single stars object for each date (2021-02-07 and 2021-02-16)

combined_lights_2021_02_07 <- st_mosaic(dnb_day38_tile85_2021039064328, 
                                        dnb_day38_tile86_2021039064329)

combined_lights_2021_02_16 <- st_mosaic(dnb_day47_tile85_2021048091106, 
                                        dnb_day47_tile86_2021048091105)

# Delete the previously loaded stars objects from environment

rm(dnb_day38_tile85_2021039064328, 
   dnb_day38_tile86_2021039064329,
   dnb_day47_tile85_2021048091106, 
   dnb_day47_tile86_2021048091105)
```

```{r}
# Visual check of combined lights outputs
plot(combined_lights_2021_02_07)
plot(combined_lights_2021_02_16)
```

## Create a blackout mask

Subtract the night lights intensity after the storm from the night lights intensity before the storm to get the difference in night lights intensity caused (presumably) by the storm. Reclassify the difference raster. Assume that any location that experienced a drop of more than 200 nW cm-2 sr-1 is a location of a blackout. 
```{r}
difference_light <- combined_lights_2021_02_16 - combined_lights_2021_02_07 > 200

plot(difference_light)
```

Assign NA to all values that have experienced a drop of less than 200 nW cm-2 sr-1.
```{r}
# change FALSE values to NA's
difference_light[difference_light == F] <- NA
```

## Vectorize the mask

```{r}
# vectorize blackout mask raster
blackout <- st_as_sf(difference_light)
```

```{r}
# fix "broken" vector dataset (i.e., fix polygons with invalid geometries)
blackout_mask <- st_make_valid(blackout)
```

## Crop the vectorized map to our ROI

Our region of interest (ROI) is the metropolitan Houston area. 
We define the ROI with a bounding box of:
-96.5, 29
-96.5, 30.5
-94.5, 30.5
-94.5, 29

```{r}
# create ROI polygon from coordinates
ROI = st_polygon(list(rbind(c(-96.5, 29), c(-96.5, 30.5), c(-94.5, 30.5), c(-94.5, 29), c(-96.5, 29))))

```

```{r}
# convert the polygon into a simple feature collection using st_sfc()
ROI <- st_sfc(ROI)

# check CRS of night lights data
st_crs(blackout_mask) # WGS84

# assign same CRS to ROI sfc
st_crs(ROI) <- 4326 
# quick check
st_crs(ROI) #WGS84
```

```{r}
# spatial subsetting in sf
intersects <- st_intersects(blackout_mask, ROI, sparse = FALSE)
cropped_mask = blackout_mask[intersects,]
```

```{r}
# re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)
cropped_mask = st_transform(cropped_mask, crs = 3083)

# quick check
st_crs(cropped_mask) #NAD83
```


## Roads Data

During normal times, highways account for a large portion of the night lights observable from space. To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas areas near highways.

Load GeoPackage containing the subset of roads that intersect the Houston metropolitan area.

```{r}
# load roads data, and isolate highways specifically
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
```

```{r}
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query)
```

```{r}
# check crs projection of highways dataset
st_crs(highways) #WGS84

# change crs to NAD83
highways = st_transform(highways, crs = 3083)
```

```{r}
# identify areas 200 m either side of all highways

st_buffer()


# dissolve buffers

st_union()


```



## Buildings Data


## Census Tract Data










